### 2018-02-22 thesis/contents/discussions.tex
Updates from ShareLaTeX

\item We are the first to image the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography. Prior to this work, CN V tractography was limited to single-tensor DTI models and cannot be delineated into the pons. Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increase the range of trigeminal-related pathology, for example MS-TN, that can be studied with neuroimaging.

\item The study, for the first time, clearly differentiated TN and MS-TN by sub-segment diffusivity differences, between cistern (TN) and brainstem peri-lesional (MS-TN) segments. This distinction in the trigeminal REZ between these two conditions have been suspected, but never confirmed. The study is the first to suggest that disruptions along the CN V in different locations may trigger similar pain symptoms.

\item We overcame the major technical barrier to detailed diffusion tractography study of the trigeminal pathway in a large population group, by developing and validate the Selective Automated Group Integrated Tractography (SAGIT) software platform. In the process, we determined the optimal T1-DWI co-registration methodology with clinical DWI sequences. Prior to this project, the co-registration method has never been exhaustively determined and reported in literature.

\item We are the first to perform merged group tractography study of the trigeminal nerve, and its primary sensory pathway, and the first to conduct such study in TN, and the first to perform Gaussian Process machine learning to the study of TN. We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in its cross-sectional detail. We discovered that the unaffected CN V diffusivity differs from controls as well as the affected side and that the affected TPT pathway and S1 pathway diffusivity differs from controls.

@@

The primary motivation for the study of the trigeminal nerve with tractography was to overcome the significant limitation of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus. Prior to the use of tractography, the structure of the trigeminal nerve could only be inferred from direct anatomical studies or postmortem dissections.

The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus was encouraging. It signified that for the first time we could segment the pontine trigeminal anatomy reliably with computer aid. Previously its identification was done by eye, and sometimes with simple heuristics based on distance from surrounding landmarks. Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be assessed.

@@

Resolving cross-fibers of the cerebral white matter was a fundamental advancement in diffusion tractography. The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging. 
A widely adopted technique was probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map. This kind of image-based probabilistic tractography was useful for whole-brain white matter studies. However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways. Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition. The acquisition time of these sequences was impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development. SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions. This meant that the methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.

These key algorithmic developments decidedly brought cross-fiber tractography in clinical neuroimaging to the realm of possibility. This is significant as the vast majority of imaging of neuropathologies in patients only exist as clinical acquisitions. We took advantage of this opportunity and explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST. The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also that the trigeminal nerve has a number of unique properties as a tractography target: 1) delineation quality is easy to judge due to its simple morphology. 2) it situates in both single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.

The ability to segment the entire trigeminal nerve also opened the avenue to measure its tissue diffusivity from tractography directly. Moreover, this method can extend to any scalar images that can be registered to the DWI space. These include specific tissue-metrics such as myelin, G-ratio, and PET images. This extended these methods to facilitate disease studies well beyond TN. We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed as never before, detailed diffusivity changes as the nerve transited from pontine tissue to the cistern. The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure. In theory the number of divisions could be infinitely large, given that a centroid curve was identified. In practice the cross-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered. Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF. White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chaism. A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research. Turns out that the merged tractography strategy simplified this task by allowed a single curve to be defined manually in the template tractography space. This greatly reduced the amount of labour for researchers, as centroids no longer need to be defined repeatedly for each subject. Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.

>>>>

\item Imaging of the brainstem segment of the CN V \textit{in vivo} using multi-tensor diffusion tractography. Prior to this work, CN V tractography was limited to single-tensor DTI models and trigeminal fibers could not be visualized as they enter into the pons. Clear delineation of the pontine CN V nerve allows reliable measurements of CN V diffusivity, therefore greatly increasing the range of study of trigeminal-related pathology that can be studied with neuroimaging, such as MS-TN.

\item Differentiation between TN and MS-TN by examining sub-segment diffusivities, and distinguishing between cistern (TN) and brainstem peri-lesional (MS-TN) segments. This pathophysiological distinction between these two entities had not been previously demonstrated using neuroimaging tools, nor had it been correlated with the clinical expression of pain.

\item Overcoming the major technical barrier to diffusion tractography study of the trigeminal pathway in a large population group, by developing and validating the Selective Automated Group Integrated Tractography (SAGIT) software platform. For the process, we determined the optimal T1-DWI co-registration methodology with clinical DWI sequences. Prior to this project, the optimal co-registration method had not been exhaustively studied.

\item Merged group tractography study of the trigeminal nerve and its primary sensory pathway as a novel use of Gaussian Process machine learning to the study of TN. We studied the trigeminal primary (CN V), secondary (trigeminopontothalamic, TPT), and tertiary (trigeminocortical, S1) white patter pathways in cross-sectional detail. Our report defines distinct bilateral changes in trigeminal neuralgia subjects, as well as changes between affected and unaffected sides.

@@

The study of trigeminal neuralgia has revolutionarized the approach to the investigation of this unique pain disorder. Additionally, an important motivation for the use of tractography to study of the trigeminal nerve was the ability to overcome the significant limitations of current conventional MR imaging in the visualization of the brainstem trigeminal nerve and the area of the trigeminal nucleus. Prior to the use of tractography, the fibers of the trigeminal nerve and connected pathways could only be inferred from anatomical studies or postmortem dissections.

The ability to resolve the pontine trigeminal nerve into the trigeminal nucleus signified that for the first time we could use software methods to segment the pontine trigeminal anatomy in a reliable manner. Prior to this, the ROI identification was done visually, and sometimes with simple heuristics methods based on distance from surrounding landmarks. Such manual methods were highly subjective and became problematic with the presence of pathology, where distortions to the nerve could not be always assessed.

@@

Diffusion tensor imaging revolutionarized how we examine brain white matter fibers, yet resolving cross-fibers of the cerebral white matter is a limitation of diffusion tractography. The brain white matter was believed to consists of 90\% complex crossing pathways, and the inability for DTI to reliably image these structure posed a fundamental challenge for neuroimaging. 
Techniques to improve visualization of white matter fibers consist in probabilistic tractography, which used tissue priors to sample a large number of directions at each step stochastically, and combined the final result as a volumetric probability map. This kind of image-based probabilistic tractography was useful for whole-brain white matter studies. However, it was constrained to the resolution of the diffusion acquisition, which was insufficient for localized and small diameter pathways. Moreover, the algorithm depended on pre-generated tissue-priors, which excluded cranial nerves. 
Deterministic algorithms such as diffusion spectral imaging (DSI), and the related multi-shell acquisition strategies attempted to model the fiber orientation distribution by dramatically increasing the number of gradient directions (256--512 directions were required) in the diffusion acquisition. The acquisition time of these sequences was impractical for the clinical environment. Reduction in acquisition time for these algorithms is currently an active area of research \cite{}.
The family of spherical harmonics (SH) based methods, which include Q-ball and constrained spherical deconvolution (CSD), were a significant development. SH-based methods modelled the fiber orientation distribution as a decomposition of SH basis functions. This meant that these methods are more robust to noise, and therefore require less number of gradient acquisitions. CSD proved to be particularly robust to noise, and can be deployed on acquisitions with only 55 directions. Another family of tractography (XST) algorithm constrained its model assumption to two fiber crossings, also proved to provide good results with 50 gradient directions.

These algorithmic developments brought imaging of crossing-fibers using tractography to the realm of possibility. We explored the feasibility of imaging the pontine trigeminal nerve with CSD and XST. The trigeminal nerve is an ideal imaging target, both due to our immediate interest in TN, and also since the trigeminal nerve has a number of unique properties as a tractography target: 1) Its immediate morphology can be easily assessed in conventional MR images. 2) it courses through single-fiber (cistern) and crossing-fiber environments; 3) it is the largest cranial nerve, and therefore a good benchmark for imaging small anatomy.

The segmentation of the subregions of the trigeminal pathway allowed for the possibility of measuring the respective diffusivities. Moreover, this method can extend to any scalar images that can be registered to the DWI space. These include specific tissue-metrics such as myelin, G-ratio, and PET images. This extended these methods to facilitate disease studies well beyond TN. We were able to divide the trigeminal nerve into 30 discrete cross-sections along its path, and observed detailed diffusivity changes as the nerve transitioned from pontine tissue to the cistern. The number of divisions chosen was arbitrary, with consideration that the number 30 could easily be divided into 3 sub-segments to represent the beginning, middle, and end of an elongated structure. In theory the number of divisions could be infinitely large, given that a centroid curve was identified. In practice the cross-sectional volume decreased with more divisions, and number of points that were sampled by each cross-sectional volume needed to be considered. Determining a common centroid curve from many subjects was computationally non-trivial, especially when the streamline lengths were highly variable due to the noise in the cistern CSF. White matter pathways are highly complex, with possible multi-directional forks, bends, and convergent pathways across a wide range of scales, some examples include the forniceal crura, zona incerta, and optic chaism. A generalized algorithm that can reliably calculate a centroid for these structure is still a topic of research. Turns out that the merged tractography strategy simplified this task by allowed a single curve to be defined manually in the template tractography space. This greatly reduced the amount of labour for researchers, as centroids no longer need to be defined repeatedly for each subject. Manually curve definition by experts can leverage known techniques in computer graphics to produce centroid for very complex structures, thereby extending SAGIT for obtaining along-the-track measure for other highly variable streamline bundles.
